---
title: "The Leslie Matrix - revisited"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

A population of fish we're interested in has four life stages: eggs (E), fry (F), juvenile (J), breeding adult (A). You are told that annually:

- Each breeding adult will produce, on average, 600 eggs
- 5% of eggs will survive to become fry
- 20% of fry will survive to become juveniles
- 16% of juveniles will survive to become adults
- Adult survival rate year to year is 90%

Recall the Leslie matrix from EDS 212. Matrix multiplication (using the dot product `%*%` of the matrix with the populations at each life stage) yields the population structure at the following year. 

### Create the Leslie matrix:

```{r}
# Make the Leslie Matrix:
fish_leslie <- matrix(c(0, 0, 0, 600, 0.05, 0, 0, 0, 0, 0.20, 0, 0, 0, 0, 0.16, 0.9), nrow = 4, ncol = 4, byrow = TRUE)

# Check it out: 
fish_leslie
```

The initial population structure is given by: 
```{r}
# Initial population structure (Roe, Fry, Juvenile, Adult): 
fish_year0 <- c(5000, 8000, 600, 400)
```


Add line-by-line comments wheres prompted in the code below, explaining what each is doing. Make sure that you are talking with your neighbors about what is really being created at each step. 

```{r}
# Model projections

# ------ WHAT HAPPENS HERE? ------ #
# ------ comment line-by-line ---- #

# The line below...
time_yr <- seq(from = 0, to = 8, by = 1)

# creates a vector with the period range for which we want to model fish population

# The line below...
# --- creates a storage matrix with the columns = fish_year0 values ; rows = projection year
proj_year <- matrix(ncol = length(fish_year0), nrow = length(time_yr))

# The line below...
# --- populates the 1st row of the storage matrix with the initial condition in year 0
# --- the rest of the rows will be set to NA at this step
proj_year[1, ] <- fish_year0



# The for loop below...
# --- populates the remaining rows of the matrix [2:8] using dotproduct between fish_leslie matrix and the vector with the fish population values from the previous cycle
# --- the first row has already been populated in the previous step
for (i in 2:length(time_yr)) {
  # --- proj_year[i-1,] returns the vector with the values of the fish population from the previous cycle
  proj_year[i,] <- fish_leslie %*% proj_year[i-1,]
}

# The rest is wrangling & visualization (run to check it out):
# --- assign the column names to the proj_year matrix
colnames(proj_year) <- c("eggs", "fry", "juvenile", "adult")

# --- convert data in the format friendly to plot different categories on a single chart (here: lifestages)
proj_df <- data.frame(time_yr, proj_year) %>% 
  # --- pivot longer for all lines excluding time_yr ("-time_yr")
  # --- pivot longer allows to merge multiple columns into a single column
  pivot_longer(cols = -time_yr, names_to = "lifestage", values_to = "stage_population")

ggplot(data = proj_df, aes(x = time_yr, y = stage_population)) +
  geom_line(aes(color = lifestage)) +
  scale_y_log10()

```

Test different combinations of survival / reproduction and starting population structures to see what happens! 

### Demo pivot longer function
```{r}
demo <- data.frame(year = c(1995, 2000, 2005),
                   dogs = c(2, 4, 6),
                   cats = c(12, 14, 18))

```

```{r}
ggplot(data = demo) +
  geom_line(aes(x= year, y = cats)) +
  geom_line(aes(x= year, y = dogs)) 
```

```{r}

demo <- pivot_longer(cols = c(cats, dogs), names_to = "species", values_to = "count" )
```

